---
title: " Data Science: Capstone HarvardX - PH125.9x - CYOP - Credit risk profile predictor"
author: "Javier Dominguez Lopez"
date: "25/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

As part of the Professional Data Science Certification Program by HarvardX, students are required to complete one last course: Data Science: Capstone HarvardX - PH125.9x which is, in reality, a capstone project designed for students to be able to put in practice all the skills learned during the 8 previous courses which comprise the Program.

The capstone is divided in two projects:

The first project consists in the design, build, training and evaluation of a Machine Learning model able to make movie recommendations to users based in existing historical data of movies' ratings by users.

The second project is a Choose-Your-Own-Project where students can decide which kind of challenge they want to deal with and which methods and algorithms they're going to use to solve the problem.

In the case of this report, the CYO project consists in training an optimal algorithm which is able to assign a credit risk profile to bank customers given certain features such as the amount and duration of the credit or the education level of the client.

This document is structured as follows:

+ Introduction
+ Overview
+ Data Ingestion
+ Data pre-processing
+ Dataset exploratory analysis
+ Model build, training, testing and evaluation.
+ Conclusion


# Overview

As explained above, the target of this project is to build an algorithm able to assign a credit risk profile "good" or "bad" to clients of a bank based on both personal and financial attributes.

For this purpose a dataset composed of 1000 observations will be used to train, test and validate the algorithm.

The original dataset used in this project has been downloaded from

+ Statlog (German Credit Data) Data Set <https://archive.ics.uci.edu/ml/machine-learning-databases/statlog/german/>

The algorithm will be developed taking into account the different variables and using different approaches to determine which one provides the best performance. The resulting model should be able to predict a credit risk profile ("good" or "bad") for any new bank customer.

Accuracy will be used as the metric to evaluate algorithms' estimations although F1-scores will also be calculated for all models.

In this project 5 models will be trained and tested, resulting in an Accuracy value which will provide an idea of how good the algorithm is at estimating credit risk profiles. 

The models which will be tested are:

+ Logistic regresion
+ Decision tree
+ Random forest
+ SVM
+ kNN   

The model with the best Accuracy will be chosen as the optimal model for this project.

\pagebreak

# Methods and analysis

## Data Ingestion

This is the code which takes care of data ingestion: 

```{r data_ingestion, message = FALSE, error = FALSE,  warning = FALSE}
### Data ingestion ###
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate")
if(!require(randomForest)) install.packages("randomForest")

library(tidyverse)
library(caret)
library(data.table)
library(lubridate)
library(randomForest)

dl <- tempfile()
download.file("https://github.com/jdominguez-github/Capstone_CYOP_German_Credit_Risk/raw/master/german.data",dl)

credit <- fread(text = gsub(" ", ",", readLines(dl)),
                 col.names = c("Checking_acc_status","Duration","Credit_history","Purpose","Credit_amount","Savings_account",
                               "Current_empl_dur","Installment_rate","Personal_status_Sex","Other_debtors_guarantors",
                               "Residence_since","Property","Age","Other_installment_plans","Housing","N_credits",
                               "Job","N_dependant_people","Telephone","Foreign","Risk"))

# Split credit data set into training set (80%) and test set (20%)
set.seed(1, sample.kind="Rounding")
test_index <- createDataPartition(credit$Risk,1,0.2,list=FALSE)

credit_test <- credit[test_index, ]
credit_train <- credit[-test_index, ]
```

By executing this code chunk we'll end up with a full data set calle "credit" which is split into a training data set called "credit_train" and a test data set called "credit_test".


## Data pre-processing

The original data set (credit) uses a codification for its variable values which makes its interpretation very hard. For this reason, a process of translation will be applied so as to create an alternative data set ("credit_trans") with more descriptive column values which will be used for exploratory analysis.

This is the original codified data description as provided in the UCI site:

### Original dataset codification

* Attribute 1:  (qualitative) - Status of existing checking account
  + A11 : ... < 0 DM
  + A12 : 0 <= ... < 200 DM
  + A13 : ... >= 200 DM / salary assignments for at least 1 year
  + A14 : no checking account
* Attribute 2:  (numerical) - Duration in month

* Attribute 3:  (qualitative) - Credit history
  + A30 : no credits taken/all credits paid back duly
  + A31 : all credits at this bank paid back duly
  + A32 : existing credits paid back duly till now
  + A33 : delay in paying off in the past
  + A34 : critical account/other credits existing (not at this bank)
* Attribute 4:  (qualitative) - Purpose
  + A40 : car (new)
  + A41 : car (used)
  + A42 : furniture/equipment
  + A43 : radio/television
  + A44 : domestic appliances
  + A45 : repairs
  + A46 : education
  + A47 : (vacation - does not exist?)
  + A48 : retraining
  + A49 : business
  + A410 : others
* Attribute 5:  (numerical) - Credit amount
* Attibute 6:  (qualitative) - Savings account/bonds
  + A61 : ... < 100 DM
  + A62 : 100 <= ... < 500 DM
  + A63 : 500 <= ... < 1000 DM
  + A64 : .. >= 1000 DM
  + A65 : unknown/ no savings account
* Attribute 7:  (qualitative) - Present employment since
  + A71 : unemployed
  + A72 : ... < 1 year
  + A73 : 1  <= ... < 4 years  
  + A74 : 4  <= ... < 7 years
  + A75 : .. >= 7 years
* Attribute 8:  (numerical) - Installment rate in percentage of disposable income
* Attribute 9:  (qualitative) - Personal status and sex
  + A91 : male   : divorced/separated
  + A92 : female : divorced/separated/married
  + A93 : male   : single
  + A94 : male   : married/widowed
  + A95 : female : single
* Attribute 10: (qualitative) - Other debtors / guarantors
  + A101 : none
  + A102 : co-applicant
  + A103 : guarantor
* Attribute 11: (numerical) - Present residence since
* Attribute 12: (qualitative) - Property
  + A121 : real estate
  + A122 : if not A121 : building society savings agreement/life insurance
  + A123 : if not A121/A122 : car or other, not in attribute 6
  + A124 : unknown / no property
* Attribute 13: (numerical) - Age in years
* Attribute 14: (qualitative) - Other installment plans 
  + A141 : bank
  + A142 : stores
  + A143 : none
* Attribute 15: (qualitative) - Housing
  + A151 : rent
  + A152 : own
  + A153 : for free
* Attribute 16: (numerical) - Number of existing credits at this bank
* Attribute 17: (qualitative) - Job
  + A171 : unemployed/ unskilled  - non-resident
  + A172 : unskilled - resident
  + A173 : skilled employee / official
  + A174 : management/ self-employed/ highly qualified employee/ officer
* Attribute 18: (numerical) - Number of people being liable to provide maintenance for
* Attribute 19: (qualitative) - Telephone
  + A191 : none
  + A192 : yes, registered under the customers name
* Attribute 20: (qualitative) - foreign worker
  + A201 : yes
  + A202 : no

### Feature translation process

This is the code that carries out column codes translation:

```{r data_preprocessing, message = FALSE, error = FALSE,  warning = FALSE}
### Data pre-processing

# Translate original data set codes into something more descriptive for exploratory analysis

credit_trans <- credit

# Checking_acc_status
credit_trans[credit$Checking_acc_status == "A11"]$Checking_acc_status <- "0"
credit_trans[credit$Checking_acc_status == "A12"]$Checking_acc_status <- "0-200"
credit_trans[credit$Checking_acc_status == "A13"]$Checking_acc_status <- ">200"
credit_trans[credit$Checking_acc_status == "A14"]$Checking_acc_status <- "No account"

# Credit_history
credit_trans[credit$Credit_history == "A30"]$Credit_history <- "No credit/Duly paid at other banks"
credit_trans[credit$Credit_history == "A31"]$Credit_history <- "Duly paid at this bank"
credit_trans[credit$Credit_history == "A32"]$Credit_history <- "Existing credit duly paid at this bank"
credit_trans[credit$Credit_history == "A33"]$Credit_history <- "Payment delay in the past"
credit_trans[credit$Credit_history == "A34"]$Credit_history <- "Critical account/Credit at other banks"

# Purpose
credit_trans[credit$Purpose == "A40"]$Purpose <- "car (new)"
credit_trans[credit$Purpose == "A41"]$Purpose <- "car (used)"
credit_trans[credit$Purpose == "A42"]$Purpose <- "furniture/equipment"
credit_trans[credit$Purpose == "A43"]$Purpose <- "radio/television"
credit_trans[credit$Purpose == "A44"]$Purpose <- "domestic appliances"
credit_trans[credit$Purpose == "A45"]$Purpose <- "repairs"
credit_trans[credit$Purpose == "A46"]$Purpose <- "education"
credit_trans[credit$Purpose == "A47"]$Purpose <- "vacation"
credit_trans[credit$Purpose == "A48"]$Purpose <- "retraining"
credit_trans[credit$Purpose == "A49"]$Purpose <- "business"
credit_trans[credit$Purpose == "A410"]$Purpose <- "others"

# Savings_account
credit_trans[credit$Savings_account == "A61"]$Savings_account <- "<100"
credit_trans[credit$Savings_account == "A62"]$Savings_account <- "100-500"
credit_trans[credit$Savings_account == "A63"]$Savings_account <- "501-1000"
credit_trans[credit$Savings_account == "A64"]$Savings_account <- ">1000"
credit_trans[credit$Savings_account == "A65"]$Savings_account <- "Unknown/No account"

# Current_empl_dur
credit_trans[credit$Current_empl_dur == "A71"]$Current_empl_dur <- "Unemployed"
credit_trans[credit$Current_empl_dur == "A72"]$Current_empl_dur <- "<1y"
credit_trans[credit$Current_empl_dur == "A73"]$Current_empl_dur <- "1y-4y"
credit_trans[credit$Current_empl_dur == "A74"]$Current_empl_dur <- "4y-7y"
credit_trans[credit$Current_empl_dur == "A75"]$Current_empl_dur <- ">7y"

# Personal_status_Sex
# New Personal_status feature
credit_trans <- credit_trans %>% mutate(Personal_status="")
credit_trans <- credit_trans %>% mutate(Sex="")

credit_trans[credit$Personal_status_Sex == "A91"]$Personal_status <- "divorced/separated"
credit_trans[credit$Personal_status_Sex == "A92"]$Personal_status <- "divorced/separated/married"
credit_trans[credit$Personal_status_Sex == "A93"]$Personal_status <- "single"
credit_trans[credit$Personal_status_Sex == "A94"]$Personal_status <- "married/widowed"
credit_trans[credit$Personal_status_Sex == "A95"]$Personal_status <- "single"
# New Sex feature
credit_trans[credit$Personal_status_Sex == "A91"]$Sex <- "male"
credit_trans[credit$Personal_status_Sex == "A92"]$Sex <- "female"
credit_trans[credit$Personal_status_Sex == "A93"]$Sex <- "male"
credit_trans[credit$Personal_status_Sex == "A94"]$Sex <- "male"
credit_trans[credit$Personal_status_Sex == "A95"]$Sex <- "female"

credit_trans <- credit_trans %>% select(-Personal_status_Sex)

# Other_debtors_guarantors
credit_trans[credit$Other_debtors_guarantors == "A101"]$Other_debtors_guarantors <- "none"
credit_trans[credit$Other_debtors_guarantors == "A102"]$Other_debtors_guarantors <- "co-applicant"
credit_trans[credit$Other_debtors_guarantors == "A103"]$Other_debtors_guarantors <- "guarantor"

# Property
credit_trans[credit$Property == "A121"]$Property <- "real estate"
credit_trans[credit$Property == "A122"]$Property <- "building society savings agreement/life insurance"
credit_trans[credit$Property == "A123"]$Property <- "car or other"
credit_trans[credit$Property == "A124"]$Property <- "unknown / no property"

# Other_installment_plans
credit_trans[credit$Other_installment_plans == "A141"]$Other_installment_plans <- "bank"
credit_trans[credit$Other_installment_plans == "A142"]$Other_installment_plans <- "stores"
credit_trans[credit$Other_installment_plans == "A143"]$Other_installment_plans <- "none"

# Housing
credit_trans[credit$Housing == "A151"]$Housing <- "rent"
credit_trans[credit$Housing == "A152"]$Housing <- "own"
credit_trans[credit$Housing == "A153"]$Housing <- "for free"

# Job
credit_trans[credit$Job == "A171"]$Job <- "unemployed/unskilled - NR"
credit_trans[credit$Job == "A172"]$Job <- "unemployed/unskilled - R"
credit_trans[credit$Job == "A173"]$Job <- "skilled/official"
credit_trans[credit$Job == "A174"]$Job <- "management/self-employed/highly qualified/officer"

# Telephone
credit_trans[credit$Telephone == "A191"]$Telephone <- "no"
credit_trans[credit$Telephone == "A192"]$Telephone <- "yes"

# Foreign
credit_trans[credit$Foreign == "A201"]$Foreign <- "yes"
credit_trans[credit$Foreign == "A202"]$Foreign <- "no"

# Risk_profile
credit_trans <- credit_trans %>% mutate(Risk_profile="")

credit_trans[credit$Risk == "1"]$Risk_profile <- "good"
credit_trans[credit$Risk == "2"]$Risk_profile <- "bad"

# Convert categorical features into factor
credit_trans$Checking_acc_status <- factor(credit_trans$Checking_acc_status)
credit_trans$Credit_history <- factor(credit_trans$Credit_history)
credit_trans$Purpose <- factor(credit_trans$Purpose)
credit_trans$Savings_account  <- factor(credit_trans$Savings_account )
credit_trans$Current_empl_dur <- factor(credit_trans$Current_empl_dur)
credit_trans$Other_debtors_guarantors <- factor(credit_trans$Other_debtors_guarantors)
credit_trans$Property <- factor(credit_trans$Property)
credit_trans$Other_installment_plans <- factor(credit_trans$Other_installment_plans)
credit_trans$Housing <- factor(credit_trans$Housing)
credit_trans$Job <- factor(credit_trans$Job)
credit_trans$Telephone <- factor(credit_trans$Telephone)
credit_trans$Foreign <- factor(credit_trans$Foreign)
credit_trans$Personal_status <- factor(credit_trans$Personal_status)
credit_trans$Sex <- factor(credit_trans$Sex)

# Convert outcome features into factor
credit_trans$Risk_profile = factor(credit_trans$Risk_profile)
credit$Risk = factor(credit$Risk)
credit_test$Risk = factor(credit_test$Risk)
credit_train$Risk = factor(credit_train$Risk)
```

The original codified data set "credit" will be used for model implementation and the translated data set "credit_trans" will be used for exploratory analysis.


## Data summary

With the initial loading and pre-processing of data complete, we can now take a look at the basic structure and stats of the data.

When looking at the general structure of the original data set we can see it consists of 1,000 observations and 21 variables of which, "Risk", is the outcome and the rest are features which can be used as potential predictors.

Risk is a factor variable with 2 levels: "good" and "bad" representing the two possible credit risk profiles that can be assigned to a client and that will be predicted by the algorithm.

We can see the summary statistics as well as a sample displaying the first 10 rows of both the original credit data set and the translated "credit_trans" data set which has the same amount of rows as it's only a translated version of the original one.


### Data overview: Original "credit" data set

``` {r data_summary_credit, message = FALSE, error = FALSE,  warning = FALSE}
# Overview of original credit dataset
# Total number of rows
c("Number of users" = nrow(credit))
# General structure
str(credit)
# Summary stats
summary(credit)
# Sample of first few rows
head(credit)
```


### Data overview: Translated "credit_trans" data set

```{r data_summary_credit_trans, message = FALSE, error = FALSE,  warning = FALSE}
# Overview of the dataset with translated variables
# General structure
str(credit_trans)
# Summary stats
summary(credit_trans)
# Sample of first few rows
head(credit_trans)
```

\pagebreak

## Data exploratory analysis

### General information

Number of credit risk profiles categorized as "good" / "bad"
```{r number_credit_profiles, echo=FALSE, message = FALSE, error = FALSE,  warning = FALSE}
cont_risk <- table(credit_trans$Risk_profile)
cont_risk
```

Proportion of credit risk profiles categorized as "good"
```{r prop_good_profiles, echo=FALSE, message = FALSE, error = FALSE,  warning = FALSE}
c("Proportion of \"good\" credit risk profiles" = mean(credit_trans$Risk_profile=="good"))
```

70% of the profiles have been categorized as "good" and 30% have been categorized as "bad". To visualize this, here's a plot of the proportion of good vs bad profiles

```{r risk_barplot, message = FALSE, error = FALSE,  warning = FALSE}
# Credit risk distribution
cont_succ <- table(credit_trans$Risk_profile)
data.frame(cont_succ) %>% 
  ggplot(aes(x= reorder(Var1,-Freq),Freq,fill=Var1)) +
  geom_bar(stat ="identity") +
  xlab("Risk") +
  ggtitle("Credit risk distribution")
```

\pagebreak

Number of male/female customers

```{r number_sex, echo=FALSE, message = FALSE, error = FALSE,  warning = FALSE}
cont_sex <- table(credit_trans$Sex)
cont_sex
```

```{r sex_barplot, message = FALSE, error = FALSE,  warning = FALSE}
# Sex distribution
cont_succ <- table(credit_trans$Sex)
data.frame(cont_succ) %>% 
  ggplot(aes(x= reorder(Var1,-Freq),Freq,fill=Var1)) +
  geom_bar(stat ="identity") +
  xlab("Sex") +
  ggtitle("Sex distribution")
```

The number of male credit clients is approximately double than the number of female credit clients.

\pagebreak

```{r job_barplot, message = FALSE, error = FALSE,  warning = FALSE}
# Job distribution
cont_succ <- table(credit_trans$Job)
data.frame(cont_succ) %>% 
  ggplot(aes(x= reorder(Var1,-Freq),Freq,fill=Var1)) +
  geom_bar(stat ="identity") +
  xlab("Job") +
  ggtitle("Job distribution") +
  theme(axis.text.x = element_text(angle = 90))
```

\pagebreak

```{r housing_barplot, message = FALSE, error = FALSE,  warning = FALSE}
# Housing distribution
cont_succ <- table(credit_trans$Housing)
data.frame(cont_succ) %>% 
  ggplot(aes(x= reorder(Var1,-Freq),Freq,fill=Var1)) +
  geom_bar(stat ="identity") +
  xlab("Housing") +
  ggtitle("Housing distribution")
```

Most credit clients are skilled workers.

\pagebreak

```{r svg_acc_barplot, message = FALSE, error = FALSE,  warning = FALSE}
# Savings account distribution
cont_succ <- table(credit_trans$Savings_account)
data.frame(cont_succ) %>% 
  ggplot(aes(x= reorder(Var1,-Freq),Freq,fill=Var1)) +
  geom_bar(stat ="identity") +
  xlab("Savings account") +
  ggtitle("Savings account distribution") +
  theme(axis.text.x = element_text(angle = 90))
```

A majority of credit clients are owners of their house

\pagebreak

```{r purpose_barplot, message = FALSE, error = FALSE,  warning = FALSE}
# Purpose distribution
cont_succ <- table(credit_trans$Purpose)
data.frame(cont_succ) %>% 
  ggplot(aes(x= reorder(Var1,-Freq),Freq,fill=Var1)) +
  geom_bar(stat ="identity") +
  xlab("Purpose") +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Purpose distribution")
```

\pagebreak

```{r age_dist, message = FALSE, error = FALSE,  warning = FALSE}
# Age distribution
credit_trans %>% ggplot(aes(Age)) + 
  geom_histogram(aes(y=..density..),col="red",fill="blue",alpha=.2) +
  ggtitle("Age distribution") + 
  geom_vline(aes(xintercept=mean(Age)),color="blue", linetype="dashed", size=1)+
  geom_density(alpha=.2) 
```

\pagebreak

```{r amount_dist, message = FALSE, error = FALSE,  warning = FALSE}
# Credit amount distribution
credit_trans %>% ggplot(aes(Credit_amount)) + 
  geom_histogram(aes(y=..density..),col="red",fill="red",alpha=.2) +
  ggtitle("Credit amount distribution") + 
  geom_vline(aes(xintercept=mean(Credit_amount)),color="blue", linetype="dashed", size=1)+
  geom_density(alpha=.2) 
```

\pagebreak

```{r duration_dist, message = FALSE, error = FALSE,  warning = FALSE}
# Duration distribution
credit_trans %>% ggplot(aes(Duration)) + 
  geom_histogram(aes(y=..density..),col="red",fill="green",alpha=.2) +
  ggtitle("Duration distribution") + 
  geom_vline(aes(xintercept=mean(Duration)),color="blue", linetype="dashed", size=1)+
  geom_density(alpha=.2) 
```

\pagebreak

Features correlation check

```{r age_barplot, message = FALSE, error = FALSE,  warning = FALSE}
# Age boxplot
credit_trans %>% ggplot(aes(Risk_profile,Age,fill=Risk_profile,alpha=.5)) + 
  geom_boxplot() +
  ggtitle("Age boxplot") +
  xlab("Risk profile")
```

\pagebreak

```{r age_lines, message = FALSE, error = FALSE,  warning = FALSE}
# Quality of profiles by Age
credit_trans %>% 
  group_by(Age) %>% 
  summarize(avg = mean(Risk_profile=="good")) %>%
  ggplot(aes(Age,avg)) +
  geom_point() +
  geom_smooth() +
  ggtitle("Quality of profiles by Age") +
  ylab("Average of \"good\" risk profiles")
```

The two plots above show there's a correlation between age and risk profile quality. The older the client the higher the chance for them to have a good risk profile. One could expect this as we can make the assumption that older clients will have a more stable economic situation than younger ones.

\pagebreak


```{r amount_barplot, message = FALSE, error = FALSE,  warning = FALSE}
# Credit amount boxplot
credit_trans %>% ggplot(aes(Risk_profile,Credit_amount,fill=Risk_profile,alpha=.5)) + 
  geom_boxplot() +
  ggtitle("Credit amount boxplot") +
  xlab("Risk profile")
```

\pagebreak

```{r amount_bars, message = FALSE, error = FALSE,  warning = FALSE}
# Quality of profiles by Credit_amount (bars)
credit_trans %>% 
  mutate(amount_1k = round(Credit_amount/1000)) %>% 
  group_by(amount_1k) %>% 
  summarize(avg_good = mean(Risk_profile=="good")) %>% 
  ggplot(aes(reorder(amount_1k,-avg_good),avg_good,fill=amount_1k)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Credit amount (in k DM)") +
  ggtitle("Quality of profiles by Credit_amount (bars)") +
  ylab("Average of \"good\" risk profiles")
```

\pagebreak

```{r amount_lines, message = FALSE, error = FALSE,  warning = FALSE}
# Quality of profiles by Credit_amount (line)
credit_trans %>% 
  mutate(amount_1k = round(Credit_amount/1000)) %>% 
  group_by(amount_1k) %>% 
  summarize(avg = mean(Risk_profile=="good")) %>%
  ggplot(aes(amount_1k,avg)) +
  geom_point() +
  geom_smooth() +
  xlab("Credit amount (in k DM)") +
  ggtitle("Quality of profiles by Credit_amount (line)") +
  ylab("Average of \"good\" risk profiles")
```

As shown in the three previous plots, there seems to be a trend by which holders of higher amount credits have a worse risk profile than lower ones. This makes sense as high amount credits are riskier than lower amount ones.

\pagebreak


```{r duration_barplot, message = FALSE, error = FALSE,  warning = FALSE}
# Duration boxplot
credit_trans %>% ggplot(aes(Risk_profile,Duration,fill=Risk_profile,alpha=.5)) + 
  geom_boxplot() +
  ggtitle("Duration boxplot") +
  xlab("Risk profile")
```

```{r duration_bars, message = FALSE, error = FALSE,  warning = FALSE}
# Quality of profiles by Duration (bars)
credit_trans %>% 
  group_by(Duration) %>% 
  summarize(avg_good = mean(Risk_profile=="good")) %>% 
  ggplot(aes(reorder(Duration,-avg_good),avg_good,fill=Duration)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Duration") +
  ggtitle("Quality of profiles by Duration (bars)") +
  ylab("Average of \"good\" risk profiles")
```

```{r duration_lines, message = FALSE, error = FALSE,  warning = FALSE}
# Quality of profiles by Duration (lines)
credit_trans %>% 
  group_by(Duration) %>% 
  summarize(avg = mean(Risk_profile=="good")) %>%
  ggplot(aes(Duration,avg)) +
  geom_point() +
  geom_smooth() +
  ggtitle("Quality of profiles by Duration (lines)") +
  ylab("Average of \"good\" risk profiles")
```

The three plots above show that clients with longer duration credits have a worse risk profile than shorter ones. This makes sense as the risk is higher when the bank has to face the possibility of default for a longer time. 


\pagebreak

```{r quality_by_sex, message = FALSE, error = FALSE,  warning = FALSE}
# Quality of profiles by Sex
credit_trans %>% 
  group_by(Sex) %>% 
  summarize(avg_good = mean(Risk_profile=="good")) %>% 
  ggplot(aes(reorder(Sex,-avg_good),avg_good,fill=Sex)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Sex") +
  ggtitle("Quality of profiles by Sex") +
  ylab("Average of \"good\" risk profiles")
```

\pagebreak

```{r quality_by_job, message = FALSE, error = FALSE,  warning = FALSE}
# Quality of profiles by Job
credit_trans %>% 
  group_by(Job) %>% 
  summarize(avg_good = mean(Risk_profile=="good")) %>% 
  ggplot(aes(reorder(Job,-avg_good),avg_good,fill=Job)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Job") +
  ggtitle("Quality of profiles by Job") +
  ylab("Average of \"good\" risk profiles")
```

\pagebreak

```{r quality_by_housing, message = FALSE, error = FALSE,  warning = FALSE}
# Quality of profiles by Housing
credit_trans %>% 
  group_by(Housing) %>% 
  summarize(avg_good = mean(Risk_profile=="good")) %>% 
  ggplot(aes(reorder(Housing,-avg_good),avg_good,fill=Housing)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Housing") +
  ggtitle("Quality of profiles by Housing") +
  ylab("Average of \"good\" risk profiles")
```

\pagebreak

```{r quality_by_savings_account, message = FALSE, error = FALSE,  warning = FALSE}
# Quality of profiles by Savings_account
credit_trans %>% 
  group_by(Savings_account) %>% 
  summarize(avg_good = mean(Risk_profile=="good")) %>% 
  ggplot(aes(reorder(Savings_account,-avg_good),avg_good,fill=Savings_account)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Savings account") +
  ggtitle("Quality of profiles by Savings_account") +
  ylab("Average of \"good\" risk profiles")
```


\pagebreak

```{r quality_by_purpose, message = FALSE, error = FALSE,  warning = FALSE}
# Quality of profiles by Purpose
credit_trans %>% 
  group_by(Purpose) %>% 
  summarize(avg_good = mean(Risk_profile=="good")) %>% 
  ggplot(aes(reorder(Purpose,-avg_good),avg_good,fill=Purpose)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Purpose") +
  ggtitle("Quality of profiles by Purpose") +
  ylab("Average of \"good\" risk profiles")
```

Sex, Job, Housing, Savings account and Purpose seem to have a less significant correlation with the risk profile.

\pagebreak

## Modeling approach

The data frame perf_results is created to keep track of the different models performance so as they can be easily compared.

```{r results_df, message = FALSE, error = FALSE,  warning = FALSE}
perf_results <- data_frame()
```

Also, a function for F1-scores calculation is created as it will be applied to all models.

The F1-score metric is calculated as:

$$ F_{1}-Score  =  2 * \frac{precision*recall}{precision+recall} $$

```{r F1_fun, message = FALSE, error = FALSE,  warning = FALSE}
# F1-score calculation
f1 <- function(y_hat,y){
  precision <- posPredValue(y_hat, y, positive="1")
  recall <- sensitivity(y_hat, y, positive="1")
  F1 <- (2 * precision * recall) / (precision + recall)
  F1
}
```


All models are trained and parameters fitted by using the functions included in the caret package.


### Model 1: Logistic regresion

Logistic regression is an algorithm used very commonly for binary classification problems although it can be applied in many other cases. 

In this case, considering the apparent high correlation between some predictors and the outcome, the first approach will be to apply a logistic regression model to the data.

All variables will be considered as predictors for the model fitting process.

```{r glm, message = FALSE, error = FALSE,  warning = FALSE}
#------------------------
### Logistic regresion
#------------------------

# trainctrl <- trainControl(verboseIter = TRUE)

# Train model
fit_glm <- train(Risk ~ ., method="glm", data = credit_train)

# Calculate predictions using fitted model
y_hat_glm <- predict(fit_glm, credit_test, type = "raw")

# Display results
cm_glm <- confusionMatrix(y_hat_glm,credit_test$Risk)
Acc_glm <- cm_glm$overall[["Accuracy"]]
F1_glm <- f1(y_hat_glm,credit_test$Risk)

# Save first metric result in perf_results
perf_results <- data_frame(method = "Logistic regresion", Accuracy = Acc_glm, F1_score = F1_glm)
perf_results %>% knitr::kable()
```

\pagebreak

### Model 2: Decision tree

Parameter *cp* will be optimized with cross-validation.

```{r dt1, message = FALSE, error = FALSE,  warning = FALSE}
#------------------------
### Decision tree
#------------------------

# Train model
fit_dt <- train(Risk ~ ., data = credit_train,method="rpart",
                   tuneGrid = data.frame(cp = seq(0, 0.05, 0.002)), 
                trControl = trainControl(method = "cv"))

# Calculate predictions using fitted model and check results
y_hat_dt <- predict(fit_dt, credit_test, type = "raw")
cm_dt <- confusionMatrix(y_hat_dt, credit_test$Risk)
Acc_dt <- cm_dt$overall[["Accuracy"]]
F1_dt <- f1(y_hat_dt,credit_test$Risk)
```

Optimal *cp* parameter for Decision tree

```{r dt2, message = FALSE, error = FALSE,  warning = FALSE}
# Optimal cp parameter
ggplot(fit_dt)
fit_dt$bestTune
```

Final model for decision tree

```{r dt3, message = FALSE, error = FALSE,  warning = FALSE}
# Tree visualization
plot(fit_dt$finalModel, margin = 0.1)
text(fit_dt$finalModel, cex = 0.75)
```

Decision tree results

```{r dt4, message = FALSE, error = FALSE,  warning = FALSE}
# Save metric in perf_results
perf_results <- bind_rows(perf_results, data_frame(method="Decision tree", Accuracy = Acc_dt, F1_score = F1_dt ))
perf_results %>% knitr::kable()
```

\pagebreak

### Model 3: Random forest

Let's see if bagging multiple decision trees by using Random forest can improve the previous result.

Parameter *mtry* will be optimized by using cross-validation with the train() function. Parameter *ntree* will be set at a fixed value of 1000.

```{r rf1, message = FALSE, error = FALSE,  warning = FALSE}
# Warning: Please note that this code takes a substantial amount of time to execute

trainctrl <- trainControl(number = 10)

# Parameters (mtry) fit. ntree parameter is set at a fixed value of 1000
fit_rf <- train(Risk ~ .
                , data = credit_train,method="rf",
                tuneGrid = data.frame(mtry = seq(1,15)),ntree=1000, trControl = trainctrl)
```

Optimal *mtry* parameter for Random forest

```{r rf2, message = FALSE, error = FALSE,  warning = FALSE}
# Optimal mtry parameter
print(fit_rf)
ggplot(fit_rf)
fit_rf$bestTune
```

Feature importance analysis

```{r rf3, message = FALSE, error = FALSE,  warning = FALSE}
# Feature importance analysis
imp <- varImp(fit_rf)
imp
```


Random forest results

```{r rf4, message = FALSE, error = FALSE,  warning = FALSE}
# Calculate predictions using fitted model and check results
y_hat_rf <- predict(fit_rf, credit_test, type = "raw")
cm_rf <- confusionMatrix(y_hat_rf,credit_test$Risk)
Acc_rf <- cm_rf$overall[["Accuracy"]]
F1_rf <- f1(y_hat_rf,credit_test$Risk)

# Save metric in perf_results
perf_results <- bind_rows(perf_results, data_frame(method="Random forest", Accuracy = Acc_rf, F1_score = F1_rf ))
perf_results %>% knitr::kable()
```

\pagebreak

### Model 4: SVM

```{r svm1, message = FALSE, error = FALSE,  warning = FALSE}
#------------------------
### SVM with Linear Kernel
#------------------------

# Set up Repeated k-fold Cross Validation
train_control <- trainControl(method="repeatedcv", number=25, repeats=3)

# Fit the model 
svm <- train(Risk ~ ., data = credit_train, 
              method = "svmLinear", trControl = train_control)
```

View of the SVM model

```{r svm2, message = FALSE, error = FALSE,  warning = FALSE}
#View the model
svm

```

SVM results

```{r svm3, message = FALSE, error = FALSE,  warning = FALSE}
# Calculate predictions using fitted model and check results
y_hat_svm <- predict(svm, credit_test, type = "raw")
cm_svm <- confusionMatrix(y_hat_svm, credit_test$Risk)
Acc_svm <- cm_svm$overall[["Accuracy"]]
F1_svm <- f1(y_hat_svm,credit_test$Risk)

# Save metric in perf_results
perf_results <- bind_rows(perf_results, data_frame(method="SVM with Linear Kernel", Accuracy = Acc_svm, F1_score = F1_svm ))
perf_results %>% knitr::kable()
```

\pagebreak

### Model 5: kNN

Considering the relatively high number of variables, kNN is not expected to outperform the other methods.

```{r knn1, message = FALSE, error = FALSE,  warning = FALSE}
control <- trainControl(method = "cv", number = 10, p = .9)

# Fit the model 
fit_knn <- train(Risk ~ ., method = "knn", 
                      data = credit_train,
                      tuneGrid = data.frame(k = seq(3, 61, 2)))
```

Optimal K parameter

```{r knn2, message = FALSE, error = FALSE,  warning = FALSE}
# Optimal K parameter
ggplot(fit_knn)
fit_knn$bestTune

```

**kNN results**

```{r knn3, message = FALSE, error = FALSE,  warning = FALSE}
# Calculate predictions using fitted model and check results
y_hat_knn <- predict(fit_knn, credit_test, type = "raw")
cm_knn <- confusionMatrix(y_hat_knn, credit_test$Risk)
Acc_knn <- cm_knn$overall[["Accuracy"]]
F1_knn <- f1(y_hat_knn,credit_test$Risk)

# Save metric in perf_results
perf_results <- bind_rows(perf_results, data_frame(method="kNN", Accuracy = Acc_knn, F1_score = F1_knn ))
perf_results %>% knitr::kable()
```

\pagebreak

# Results

This is the final result

```{r results, message = FALSE, error = FALSE,  warning = FALSE}
perf_results %>% knitr::kable()
```

The highest value for Accuracy is

```{r highest_acc, echo=FALSE}
Acc_glm
```

provided by the **logistic regression** model.

# Conclusion

A model to predict credit risk profile qualities has been built by testing different approaches and choosing the one with best results.

The optimal model accounts for the variability due to the different features available in the dataset.

The final accuracy obtained with the optimal model, which turned out to be **logistic regression**, is

```{r final_acc, echo=FALSE}
Acc_glm
```




